---
title: "Assets to Portfolio Returns"
output: html_notebook
---
```{r setup, message = FALSE, include = FALSE}
# packages required for this post
for (pkg in c('tidyverse', 'tidyquant', 'timetk', 'tibbletime', 'highcharter')) 
  if (!requireNamespace(pkg)) install.packages(pkg)
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

In a [previous post](http://www.reproduciblefinance.com/2017/09/25/asset-prices-to-log-returns/), we reviewed how to import daily prices and 4 methods to transform daily prices to monthly log returns, and then we [visualized those individual asset returns]()http://www.reproduciblefinance.com/2017/09/28/visualizing-asset-returns/.  Today, we will finally combine those assets into a portfolio and analyze them as a coherent mass. 

By way of quick reminder, our 5-asset portfolio will consist of the following.

    + SPY (S&P500 fund) weighted 25%
    + EFA (a non-US equities fund) weighted 25%
    + IJS (a small-cap value fund) weighted 20%
    + EEM (an emerging-mkts fund) weighted 20%
    + AGG (a bond fund) weighted 10%
    
A brief interlude the concept of a portfolio. A portfolio is a weighted collection of assets (its etymology harkens back to the Latin for "to carry sheets of paper", which I s'pose made its way to the world of investments because securities used to actually be on sheets of paper too, and a portfolio of investments involved sheets of paper representing interests in several assets).  Indeed, we could go to a more abstract level and say a portfolio is a weighted collection of returns and we use asset names as convenient labels. The theoretical reason that rational investors prefer a portfolio to a single asset is that a portfolio can offer a better risk/return tradeoff due to low or negavitve covariance amongst portfolio components. 

Back to the task at hand: build our portfolio, analyze returns over time in a few different ways, and then visualize those returns in a few different ways.

Let's load up our packages.

```{r}
library(tidyverse)
library(tidyquant)
library(timetk)
library(tibbletime)
library(highcharter)
```

To get our objects into the global environment, the next code chunk is copy/paste from the previous post: we will create one `xts` object and one `tibble`, in long/tidy format, of monthly log returns.

```{r}
# The symbols vector holds our tickers. 
symbols <- c("SPY","EFA", "IJS", "EEM","AGG")

# The prices object will hold our raw price data throughout this book.
prices <- 
  getSymbols(symbols, src = 'yahoo', from = "2005-01-01", 
             auto.assign = TRUE, warnings = FALSE) %>% 
  map(~Ad(get(.))) %>% 
  reduce(merge) %>%
  `colnames<-`(symbols)
# XTS method
prices_monthly <- to.monthly(prices, indexAt = "last", OHLC = FALSE)
asset_returns_xts <- na.omit(Return.calculate(prices_monthly, method = "log"))

# Tidyverse method, to long, tidy format
asset_returns_long <- 
  prices %>% 
  to.monthly(indexAt = "last", OHLC = FALSE) %>% 
  tk_tbl(preserve_index = TRUE, rename_index = "date") %>%
  gather(asset, returns, -date) %>% 
  group_by(asset) %>%  
  mutate(returns = (log(returns) - log(lag(returns))))
```


## To the Portfolio Station

Now on to constructing a portfolio and calculating volatility. To turn these 5 ETFs in a porftolio we need to assign them weights. Let's first create a weights vector.


```{r, message = FALSE}
w <- c(0.25, 0.25, 0.20, 0.20, 0.10)
```

Before we use the weights in our calculations, a quick sanity check in the next code chunk. This might not be necessary with 5 assets as we have today, but good practice because if we had 50 assets it could save us a lot of grief to catch a mistake early.

```{r Weights Sanity Check}
# Make sure the weights line up with assets.
asset_weights_sanity_check <- tibble(w, symbols)
asset_weights_sanity_check
```

Does that tibble match up with the portfolio we want to create? Looks good to me.

Finally, make sure the weights sum to 100%, or 1. Again, we can eyeball this with 5 assets, but with 50 assets it would be easier to run the sanity check.

```{r}
sum(asset_weights_sanity_check$w)
```

They sum to 1. Good to go and on to portfolio returns.

### From Individual Returns to One Portfolio Return

We will start with the textbook equation for the return of a multi-asset portfolio which is: 
$$Return_{portfolio} = W_{1}*Return_{asset1}~+~W_{2}*Return_{asset2}~+~W_{3}*Return_{asset3}~+~W_{4}*Return_{asset4}~+~W_{5}*Return_{asset5}$$

```{r By Hand Return}

# Let's assign each asset a weight from our weights vector above.

w_1 <- w[1]
w_2 <- w[2]
w_3 <- w[3]
w_4 <- w[4]
w_5 <- w[5]

# And each asset has a return as well, stored in our 
# asset_returns_xts object.

asset1 <- asset_returns_xts[,1]
asset2 <- asset_returns_xts[,2]
asset3 <- asset_returns_xts[,3]
asset4 <- asset_returns_xts[,4]
asset5 <- asset_returns_xts[,5]

portfolio_returns_byhand <-   
  (w_1 * asset1) + 
  (w_2 * asset2) + 
  (w_3 * asset3) +
  (w_4 * asset4) + 
  (w_5 * asset5)

names(portfolio_returns_byhand) <- "returns"
```


For our first method, we will stay in the `xts` world and use the `Return.portfolio` function from the `Performance Analytics` package.  You might have noticed that we didn't explicitly load that package and that is because `tidyquant` imports this package for us.  The function requires two arguments for a portfolio, an `xts` object of returns and a vector of weights.  It's not necessary but we are also going to set `rebalance_on = "months"` so we can confirm it matches our by hand calculations above. Remember, in the by hand equation, we set the portfolio weights as fixed, meaning they never change on a month to month basis. That is equivalent to rebalancing every month. In practice, that would be quite rare. Once we confirm that it matches our by hand, we can toggle over to a more realistic annual rebalancing by chaning the argument to `rebalance_on = "years"`.


```{r}
portfolio_returns_xts_rebalanced_monthly <- 
  Return.portfolio(asset_returns_xts, weights = w, rebalance_on = "months") %>%
  `colnames<-`("returns")

```


```{r}
portfolio_returns_xts_rebalanced_yearly <- Return.portfolio(asset_returns_xts, weights = w, rebalance_on = "years") %>%
  `colnames<-`("returns")
```


```{r}
portfolio_returns_tq_rebalanced_monthly <- 
  asset_returns_long %>%
  tq_portfolio(assets_col  = asset, 
               returns_col = returns,
               weights     = w,
               col_rename  = "returns",
               rebalance_on = "months")
```


```{r}
portfolio_returns_tq_rebalanced_yearly <- 
  asset_returns_long %>%
  tq_portfolio(assets_col  = asset, 
               returns_col = returns,
               weights     = w,
               col_rename  = "returns",
               rebalance_on = "years")
```




### Visualize

First, let's use `highcharter` to  visualize the `xts` formatted returns.

Highcharter is fantastic for visualizing a time series or many time series.  First, we set `highchart(type = "stock")` to get a nice time series line. Then we add each of our series to the highcharter code flow. In this case, we'll add our columns from the xts object.

```{r}
highchart(type = "stock") %>% 
  hc_title(text = "Porftolio Monthly Returns") %>%
  hc_add_series(portfolio_returns_xts_rebalanced_yearly$returns, 
                  name = "Rebalanced Yearly", color = "cornflowerblue") %>%
  hc_add_theme(hc_theme_flat()) %>%
  hc_navigator(enabled = FALSE) %>% 
  hc_scrollbar(enabled = FALSE)

```

Let's add one of our more volatile assets and see if anything looks interesting

```{r}
portfolio_returns_xts_rebalanced_yearly$mean <- 
  round(mean(portfolio_returns_xts_rebalanced_yearly$returns), 4)
```

```{r}

highchart(type = "stock") %>% 
  hc_title(text = "Porftolio Monthly Returns") %>%
  hc_add_series(portfolio_returns_xts_rebalanced_yearly$returns, 
                  name = "Rebalanced Yearly", color = "cornflowerblue", lineWidth = 1.5) %>%
  hc_add_series(asset_returns_xts$EEM, 
                  name = "Emerging markets", color = "pink", lineWidth = 1.5) %>%
 # hc_add_series(portfolio_returns_xts_rebalanced_yearly$mean, 
  #                name = "Mean Portfolio", color = "purple", lineWidth = 1.5) %>%
  hc_add_theme(hc_theme_flat()) %>%
  hc_navigator(enabled = FALSE) %>% 
  hc_scrollbar(enabled = FALSE)

```


Take a look at the chart. It has a line for the monthly log returns of 3 of our ETFs (and in my opinion it's already starting to get crowded). We might be able to pull some useful intuition from this chart. Perhaps one of our ETFs remained stable the 2008 financial crisis, or had an era of consistently negative/positive returns. Highcharter is great for plotting time series line charts.

Highcharter does have the capacity for histogram making. One method is to first call the base function `hist` on the data along with the arguments for breaks and `plot = FALSE`. Then we can call `hchart` on that object. 

```{r}
hc_portfolio <- hist(portfolio_returns_xts_rebalanced_yearly$returns, breaks = 50, plot = FALSE)

hchart(hc_portfolio) %>% 
  hc_title(text = "Portfolio Log Returns Distribution")
```

Nothing wrong with that chart and it shows us the distribution of SPY returns. However, `highcharter` is missing an easy way to chart multiple histograms and to add density lines to those multiple histograms.  The functionality is fine for one set of returns (as we'll see when we get to the portfolio) but here we want to see the distribution of all of our returns series together. 

For that, we will head to the tidyverse and use `ggplot2` on our tidy `tibble` `assets_returns_long`. Because it is in long, tidy format, and it is grouped by the 'asset' column, we can chart the asset histograms collectively on one chart. 

```{r}
# Make so all titles centered in the upcoming ggplots
theme_update(plot.title = element_text(hjust = 0.5))

portfolio_returns_tq_rebalanced_yearly %>% 
  ggplot(aes(x = returns)) + 
  geom_histogram(alpha = 1, binwidth = .01, fill = "cornflowerblue", color = "cornflowerblue")
```

Let's try to compare the portfolio distribution to that of our assets. Use the alpha argument to make the asset histograms a bid faded, since there are more of them and the portfolio return is what we really want to see.

```{r}
portf_asset_returns_hist <- 
  ggplot(data = asset_returns_long, aes(x = returns, fill = asset)) + 
  geom_histogram(alpha = 0.15, binwidth = .01) +
  geom_histogram(data = portfolio_returns_tq_rebalanced_yearly, fill = "cornflowerblue") +
  ggtitle("Monthly Returns Since 2005")

portf_asset_returns_hist
```


Maybe we don't want to use a histogram, but instead want to use a density line to visualize the various distributions. We can use the `stat_density(geom = "line", alpha = 1)` function to do this. The `alpha` argument is selecting a line thickness. Let's also add a label to the x and y axis with the `xlab` and `ylab` functions.

```{r}

portfolio_returns_tq_rebalanced_yearly %>% 
  ggplot(aes(x = returns)) +
  stat_density(geom = "line", alpha = 1, colour = "blue") +
  ggtitle("Portfolio Monthly Returns Since 2005") +
  xlab("monthly returns") +
  ylab("distribution") 

```

That chart is quite digestible, but we can also `facet_wrap(~asset)` to break the densities out into individual charts.

```{r}
# could also combine to one for v

portf_asset_returns_density <- 
  ggplot(data =  asset_returns_long, 
         aes(x = returns)) +
  stat_density(geom = "line", 
               alpha = 1,
               aes(colour = asset)) +
  stat_density(data = portfolio_returns_tq_rebalanced_yearly, 
               geom = "line",
               alpha = 1,
               aes(colour = returns)) +
  ggtitle("Portfolio and Asset Monthly Returns Since 2005") +
  xlab("monthly returns") +
  ylab("distribution") 

portf_asset_returns_density
```

Now we can combine all of our ggplots into one nice, faceted plot. 

At the same time, to add to the aesthetic toolkit a bit, we will do some editing to the label colors. First off, let's choose a different color besides black to be the theme. I will go with cornflower blue, because it's a nice shade and I don't see it used very frequently elsewhere. Once we have a color, we can choose the different elements of the chart to change in the the `theme` function. I make a lot of changes here by way of example but feel free to comment out a few of those lines and see the different options.


```{r}
portfolio_returns_tq_rebalanced_yearly %>% 
  ggplot(aes(x = returns)) +
  geom_histogram(alpha = 1, binwidth = .01, colour = "cornflowerblue", fill = "cornflowerblue") +
  stat_density(geom = "line", alpha = 1) +
  ggtitle("Portfolio Monthly Returns Since 2005") +
  xlab("monthly returns") +
  ylab("distribution") +
  # Lots of elements can be customized in the theme() function
  theme(plot.title = element_text(colour = "cornflowerblue"), 
        axis.text.x = element_text(colour = "cornflowerblue"), 
        axis.text = element_text(colour = "cornflowerblue"), 
        axis.ticks.x = element_line(colour = "cornflowerblue"), 
        axis.text.y = element_text(colour = "cornflowerblue"), 
        axis.ticks.y = element_line(colour = "cornflowerblue"),
        axis.title = element_text(colour = "cornflowerblue")
        )
```
