---
title: "rviews-assets-to-portfolio-returns"
output:
  html_notebook: default
---
```{r setup, message = FALSE, include = FALSE}
for (pkg in c('tidyverse', 'tidyquant', 'timetk')) 
  if (!requireNamespace(pkg)) install.packages(pkg)
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

Today, we go back a bit to where we probably should have started in the first place, but it wouldn't have been as much fun. In our previous work on [volatility](https://rviews.rstudio.com/2017/07/12/introduction-to-volatility/), we zipped through the steps of data import, tidy and transformation.  Let's correct that oversight and do some spade work on transforming daily asset prices to monthly portfolio log returns.

Our 5-asset portfolio will consist of the following securities and weights:

    + SPY (S&P500 fund) weighted 25%
    + EFA (a non-US equities fund) weighted 25%
    + IJS (a small-cap value fund) weighted 20%
    + EEM (an emerging-mkts fund) weighted 20%
    + AGG (a bond fund) weighted 10%
    
A brief interlude on portfolios. A portfolio is a weighted collection of assets (its etymology harkens back to the Latin for "to carry sheets of paper", which I s'pose made its way to the world of investments because securities used to be sheets of paper).  The theoretical reason that rational investors prefer a portfolio to a single asset is that a portfolio can offer a better risk/return trade off due to low or negative covariance amongst portfolio components. 

Back to the task at hand: transform a collection of daily ETF prices into an object of portfolio log returns.

Let's load up our packages.

```{r}
library(tidyverse)
library(tidyquant)
library(timetk)
```

First, we import daily prices for the 5 ETFs, using `getSymbols` to grab the data,  `map(~Ad(get(.)))` to select adjusted prices only, and `reduce(merge)` to mash our 5 prices into one `xts` object.

```{r}
# The symbols vector holds our tickers. 
symbols <- c("SPY","EFA", "IJS", "EEM","AGG")

# The prices object will hold our raw price data throughout this book.
prices <- 
  getSymbols(symbols, src = 'yahoo', from = "2005-01-01", 
             auto.assign = TRUE, warnings = FALSE) %>% 
  map(~Ad(get(.))) %>% 
  reduce(merge) %>%
  `colnames<-`(symbols)
```

Next we convert those daily adjusted prices to monthly log returns, using two methods. For the first method, we stay in the `xts` world.

```{r}
prices_monthly <- to.monthly(prices, indexAt = "last", OHLC = FALSE)
asset_returns_xts <- na.omit(Return.calculate(prices_monthly, method = "log"))
```

For the second method, we will head to the `tidyverse/tidyquant` world. We will convert from `xts` to `tibble` using a call to `tk_tbl(preserve_index = TRUE, rename_index = "date")`. We will add a column for log returns using `mutate(returns = (log(returns) - log(lag(returns))))`. 
```{r}
# Tidyverse method, to long, tidy format
asset_returns_long <- 
  prices %>% 
  to.monthly(indexAt = "last", OHLC = FALSE) %>% 
  tk_tbl(preserve_index = TRUE, rename_index = "date") %>%
  gather(asset, returns, -date) %>% 
  group_by(asset) %>%  
  mutate(returns = (log(returns) - log(lag(returns))))
```

Have a peek at both asset return objects. 

```{r}
head(asset_returns_xts)
head(asset_returns_long)
```

Do we notice any differences?

First, have a look at the left most part of `asset_returns_xts`, where the date is stored. The `asset_returns_xts` has a date index, not a column. It is accessed via `index(asset_returns_xts)`. `asset_returns_long` has a column called "date", accessed via the `$date` convention, i.e. `asset_returns_long$date`. 

Second, notice the first date observation for January of 2005.  `asset_returns_long` contains `NA`, and `asset_returns_xts` excludes the observation completely.  Does it matter? It depends. In a few weeks when we get to the Sortino Ratio, we will see that it can matter quite a bit.

Third, `asset_returns_xts` is in wide format, which in this case means there is a column for each of our assets.  This is the format that `xts` likes and it's the format that is easier to read as a human. However, `asset_returns_long` is in long, tidy format so that each variable has its own column. It's a bit harder to read as human but the tidyverse wants data in this format.

Now on to constructing a portfolio and calculating returns. To turn these 5 ETFs into a portfolio we need to assign them weights. Let's first create a weights vector.


```{r, message = FALSE}
w <- c(0.25, 0.25, 0.20, 0.20, 0.10)
```

Before we use the weights in our calculations, a quick sanity check in the next code chunk. This might not be necessary with 5 assets as we have today, but good practice because if we had 50 assets it could save us a lot of grief to catch a mistake early.

```{r Weights Sanity Check}
# Make sure the weights line up with assets.
asset_weights_sanity_check <- tibble(w, symbols)
asset_weights_sanity_check
```

Make sure that tibble match up with the portfolio we want to create. 

Finally, make sure the weights sum to 100%, or 1. Again, we can eyeball this with 5 assets, but with 50 assets it would be easier to run the sanity check.

```{r}
sum(asset_weights_sanity_check$w)
```

They sum to 1. Good to go and on to portfolio returns.

We will start with the textbook equation for the return of a multi-asset portfolio which is: 
$$Return_{portfolio} = W_{1}*Return_{asset1}~+~W_{2}*Return_{asset2}~+~W_{3}*Return_{asset3}~+~W_{4}*Return_{asset4}~+~W_{5}*Return_{asset5}$$
Here's the Latex code for that equation. 

```{r}
# $$Return_{portfolio} = W_{1}*Return_{asset1}~+~W_{2}*Return_{asset2}~+~W_{3}*Return_{asset3}~+~W_{4}*Return_{asset4}~+~W_{5}*Return_{asset5}$$
```


We grinded through the Latex, now let's grind through the R calculation by hand instead of using built-in functions. 

First, assign each weight from our `w` vector to a variable.

Next, assign each asset return stored in  `asset_returns_xts` to a variable.

Lastly, we insert those new variables into the equation.

```{r By Hand Return}

w_1 <- w[1]
w_2 <- w[2]
w_3 <- w[3]
w_4 <- w[4]
w_5 <- w[5]


asset1 <- asset_returns_xts[,1]
asset2 <- asset_returns_xts[,2]
asset3 <- asset_returns_xts[,3]
asset4 <- asset_returns_xts[,4]
asset5 <- asset_returns_xts[,5]

portfolio_returns_byhand <-   
  (w_1 * asset1) + 
  (w_2 * asset2) + 
  (w_3 * asset3) +
  (w_4 * asset4) + 
  (w_5 * asset5)

names(portfolio_returns_byhand) <- "returns"
```


Our first portfolio returns calculation is now complete and stored as `portfolio_returns_byhand`. From a substantive perspective, we are finished and could head to visualization.  

We want to cover more methods, though, so let's head to to the `xts` world and the PerformanceAnalytics package.  We didn't explicitly load that package in the setup, because `tidyquant` imports it for us. 

We will use the`Return.portfolio` function, which requires two arguments for a portfolio, an `xts` object of asset returns and a vector of weights.  We have those at hand: `asset_returns_xts` and `w`.  It's not necessary but we will set `rebalance_on = "months"` so we can confirm it matches our by-hand calculations. Remember, in the by-hand equation, the portfolio weights are fixed, meaning they never change on a month-to-month basis. That is equivalent to re-balancing every month, which in practice would be quite rare. 


```{r}
portfolio_returns_xts_rebalanced_monthly <- 
  Return.portfolio(asset_returns_xts, weights = w, rebalance_on = "months") %>%
  `colnames<-`("returns")
```

Next let's change to a more realistic annual re-balancing and set `rebalance_on = "years"`.  This will change our results so that they no longer match our by-hand calculation, which effectively re-balanced every month (since we hard coded asset weights to be the same each month). 

```{r}
portfolio_returns_xts_rebalanced_yearly <- 
  Return.portfolio(asset_returns_xts, weights = w, rebalance_on = "years") %>%
  `colnames<-`("returns")
```

We can take a peek at our three portfolio objects and see how the annual re-balance made a small but important difference.

```{r}
head(portfolio_returns_byhand)
head(portfolio_returns_xts_rebalanced_monthly)
head(portfolio_returns_xts_rebalanced_yearly)
```

Do you notice where the annual rebalancing starts to show a difference from monthly rebalancing? 

As before, we could stop here and have accomplished our substantive task (twice already - by hand and using the built in function from PerformanceAnalytics), but we want to explore alternate methods in the world of `tidyverse/tidyquant`. We will use our long, tidy formatted `asset_returns_long` and convert to portfolio returns using the `tq_portfolio` function from `tidyquant`.  

The `tq_portfolio` function takes a `tibble` and then asks for an assets column to group by, a returns column to find return data, and a weights column. It's a wrapper for `Return.portfolio` and thus also accepts the argument `rebalance_on = "months"`.  Since we are re-balancing by months, we should again get a portfolio returns object that matches our two existing objects `portfolio_returns_byhand` and `portfolio_returns_xts_rebalanced_monthly.`   

```{r}
portfolio_returns_tq_rebalanced_monthly <- 
  asset_returns_long %>%
  tq_portfolio(assets_col  = asset, 
               returns_col = returns,
               weights     = w,
               col_rename  = "returns",
               rebalance_on = "months")

portfolio_returns_tq_rebalanced_monthly
```

If we want to re-balance annually it's the same code as above, except we set `rebalance_on = "years"`.

```{r}
portfolio_returns_tq_rebalanced_yearly <- 
  asset_returns_long %>%
  tq_portfolio(assets_col  = asset, 
               returns_col = returns,
               weights     = w,
               col_rename  = "returns",
               rebalance_on = "years")
```

We now have two more portfolio returns objects and they are both tidy tibbles. Let's take a quick look and compare how a tidy tibble of portfolio returns compares to an xts object of portfolio returns. 

```{r}
head(portfolio_returns_tq_rebalanced_yearly)
head(portfolio_returns_xts_rebalanced_yearly)
```

Again, we can see a discrepancy for January of 2005. Our `xts` object elides that date completely, while our `tibble` records it as a 0.00. 

Since there is only one column of returns, there is no wide versus long format for the `tibble` and it looks almost identical to the `xts` object.  The only difference is the date: the tibble has a column that holds the date that can be accessed with the $ operator, whereas the xts object has a date index, accessed with `index`.

That's all for today.  The `xts` and `tidyquant` object have their own uses and advantages depending on our end goal.  Next time we will think about how to visualize portfolio returns, and how the different objects fit into different visualization paradigms. 



